---
title: "Practical Machine Learning Assignment"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
In this assignment, data from the [Human Activity Recognition](http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har) is used in machine learning using a random forest model. The model is derived from a training set, which is further split into two subsets used for training and for k-fold cross validation. Then the model is applied to a test set of 20 samples and correctly predicts the outcome variable.

## Getting and Cleaning Data
First, the necessary libraries for the assignment are loaded and a seed is set for reproducibility.

```{r, message=FALSE}
library(lattice)
library(ggplot2)
library(foreach)
library(iterators)
library(caret)
library(rattle)
library(dplyr)
library(parallel)
library(doParallel)
library(pryr)
set.seed(1235)
```

The data is loaded and some exploratory analysis is done by looking at the dimensions of the data. 

```{r }
train <- read.csv("pml-training.csv", na.strings=c("NA"))
test <- read.csv("pml-testing.csv", na.strings=c("NA"))
dim(train)
dim(test)
```

## Clean the data

Check if in the observations are present NA values or missing OBS that can raise errors/bias during the model training.

```{r}
sum(complete.cases(train))
```

Too few observation to have a correct training.

### Eliminate the columns with NA/missing values

Let's see colnames

```{r}
colnames(train)
plot(colMeans(is.na(train)))
```

There are columns with a lot of missing values.

We will reatain only the columns without NA values

First covert all the data in NUMERIC form to coerce the empty factor to NA

```{r}
trainClasse = train$classe
trainRaw = train[, sapply(train, is.numeric)]
testRaw = test[, sapply(test, is.numeric)]
```

Remove columns with NA values

```{r}
trainFilter <- trainRaw[, colSums(is.na(trainRaw)) == 0]
# Attach Classe variable
trainFilter$classe = trainClasse
testFilter <- testRaw[, colSums(is.na(testRaw)) == 0]
```

Dimension

```{r}
dim(trainFilter)
dim(testFilter)
```

Removing other unuseful columns like username, timestamp and ID

```{r}
unwanted = !grepl("X|timestamp", colnames(trainFilter))
cols = colnames(trainFilter)[unwanted]
trainFilter = trainFilter %>%
  select(cols)
unwanted = !grepl("X|timestamp", colnames(testFilter))
cols = colnames(testFilter)[unwanted]
testFilter = testFilter %>%
  select(cols)
```

Get dimension of the filtered dataset

```{r}
dim(trainFilter)
dim(testFilter)
```

## Slice the data

We will slice the Training data into **Training** and **Validation** set using the 80-20 rule.

```{r}
set.seed(1202435) 
inTrain <- createDataPartition(trainFilter$classe, p=0.70, list=F)
trainData <- trainFilter[inTrain, ]
validationData <- trainFilter[-inTrain, ]
dim(trainData)
```

# Data modeling

We will fit a model using **Random Forest** for several reasons:

1. With tree-based models, **you can safely ignore** predictors correlation issues

2. Zero- and Near Zero-Variance Predictors **does not** imply on tree-based models

3. As each feature is processed separately, and the possible splits of the data don’t depend on scaling, no preprocessing like normalization or standardization of features is needed for decision tree algorithms.

## Random forest Model

```{r}
controlRf <- trainControl(method="cv", 5, allowParallel = TRUE)
modelRf <- train(classe ~ ., data=trainData, method="rf", trControl=controlRf, ntree=250)
modelRf
```

### Performance of the model on the validation data set

```{r}
predict_rf <- predict(modelRf, validationData)
confusionMatrix(validationData$classe, predict_rf)
```

Very accurate model to classify **classe** feature


# Predict Test data with RF

```{r}
resultRf <- predict(modelRf, testFilter[, -length(names(testFilter))])
resultRf
```


Finally the model predict the TEST data in the same way
